Certified Symbolic Transducer with Applications in String Solving

### 2025-03-10 Paper/main.tex
polish the paper

Even though there are various formalizations of FAs and FTs in interactive proof assistants such as Isabelle \cite{isabelle-homepage} and Coq \cite{coq-homepage}, these are predominantly based on classical definitions. However, these traditional approaches present certain limitations when applied to practical scenarios. One significant drawback is that transition labels are typically non-symbolic and finite. A conventional transition is represented as $q\xrightarrow{a}q'$, where $a$ is a character from a finite alphabet. This simplistic representation can lead to a phenomenon known as transition explosion. For example, if the alphabet $\Sigma$ encompasses the entire Unicode range, which is common in modern programming languages, it consists of $\texttt{0x10FFFF}$ distinct characters. Defining a transition from state $q$ to $q'$ that accepts any character in $\Sigma$ would require splitting into $\texttt{0x10FFFF}$ individual transitions, rendering the product of two FAs highly inefficient. Furthermore, in practical applications, it is often necessary to consider infinite alphabets, such as the set of all integers.

Symbolic Finite Automata (SFA) and Symbolic Finite-state Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic Finite Automata (FAs) and Finite-state Transducers (FTs), enhancing their applicability in practical scenarios. These symbolic models utilize transition labels defined by boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \% 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

Prior work has successfully formalized symbolic FAs within Isabelle/HOL through CertiStr \cite{cpp/KanLRS22}, demonstrating both the efficiency and effectiveness of symbolic FAs in practice. However, the formalization of FTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. FTs constitute a significantly more expressive and powerful theoretical framework compared to symbolic FAs, as evidenced by their capability to model complex string transformations such as replacement operations.
Furthermore, the seminal work of \cite{VeanesHLMB12Transducer} demonstrates the broad applicability of SFTs across diverse domains, including security-critical applications such as cross-site scripting (XSS) prevention, image transformation operations, and privacy-preserving location data processing.

@@

Leveraging the data refinement framework \cite{DBLP:conf/itp/Lammich13} formalized in Isabelle/HOL, we implement an efficient representation of states and transitions using sophisticated data structures such as hashmaps and red-black trees. Furthermore, we develop a comprehensive interval algebra that provides efficient operations for creation, verification, and manipulation of intervals, thereby facilitating the implementation of replacement operations within our string solver framework.

In summary, Our formalization makes two primary contributions. First, we achieve computational efficiency through an algebraic representation that enables compressed storage of transition labels. Second, our refinement-based formalization framework demonstrates significant extensibility, facilitating the incorporation of diverse theories of transition labels.

The paper is organized as follows.
Section \ref{sec:formalization} presents the formalization of symbolic FTs.
Section \ref{sec:product-operation} presents the product operation of symbolic FTs in the abstraction level.
Section \ref{sec_alg_refinement} presents the refinement of the product operation in Isabelle/HOL.
Section \ref{sec-app-str-solver} presents the application of SFTs in the modeling of replacement operations in string solvers.
Section \ref{sec:related-work} presents the related work.
Section \ref{sec:conclusion} concludes the paper.

@@

Symbolic Automata and Transducers \cite{cav/DAntoniV17,VeanesHLMB12Transducer,popl/VeanesHLMB12, popl/DAntoniV14, entcs/DAntoniKW18, sofsem/TammV18} represent a significant advancement in automata theory, offering improved efficiency in operations and enhanced expressiveness through algebraic theories that support infinite alphabets. This symbolic framework has been progressively extended to accommodate more complex structures: symbolic tree transducers \cite{ershov/VeanesB11} handle hierarchical data structures, while symbolic pushdown automata \cite{cav/DAntoniA14} manage nested word structures. Recent developments have further expanded the scope of symbolic techniques to include B\"{u}chi automata and omega-regular languages \cite{pacmpl/VeanesBEZ25}, enabling verification of infinite-state systems and analysis of non-terminating computations.

The efficiency, scalability, and expressive power of symbolic automata and transducers have led to their widespread adoption in numerous real-world applications. These applications span diverse domains, including: constraint solving for program analysis \cite{lpar/VeanesBM10}, security-critical sanitizer analysis for web applications \cite{uss/HooimeijerLMSV11}, runtime verification of system behaviors \cite{osdi/YaseenABCL20}, and automated program inversion for software transformation \cite{pldi/HuD17}. Each application leverages the symbolic approach's ability to handle complex patterns and infinite alphabets efficiently.

While classical automata theories have been extensively formalized in interactive theorem provers \cite{Tuerk-NFA, Lammich2014TheCA, Peter14, cpp/DoczkalKS13}, with some work on transducer formalization \cite{afp/LochmannFSTS21}, the symbolic variants of automata and transducers remain largely unexplored in formal verification. To our knowledge, CertiStr \cite{cpp/KanLRS22} represents the only existing work on symbolic automata formalization in a proof assistant. Our work advances this frontier by extending the formal treatment to both SFTs and $\varepsilon$SFAs.

A significant application of our certified transducer framework is in string constraint solving, a field that has seen intensive research development over the past decade. While our work provides formal verification guarantees, there exists a rich ecosystem of non-certified string solvers, each with distinct capabilities: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
We believe that our work will benefit the community by providing a formal foundation for string solvers.

Beyond string theories, certification efforts in SMT solving have extended to other domains. Notable among these is the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of formal verification in automated reasoning.

@@

Future work includes extending our symbolic formalization to handle variants of SFTs. One particularly promising direction is the addition of transition priorities, which would enable precise modeling of SMT-LIB's standard semantics for replacement operations.

>>>>

Even though there are various formalizations of FAs and FTs in interactive proof assistants such as Isabelle \cite{isabelle-homepage} and Coq \cite{coq-homepage}, these are predominantly based on classical definitions. However, these traditional approaches present certain limitations when applied to practical scenarios. One significant drawback is that transition labels are typically non-symbolic and finite. A conventional transition is represented as $q\xrightarrow{a}q'$, where $a$ is a character from a finite alphabet. This simplistic representation can lead to a phenomenon known as transition explosion. For example, if the alphabet $\Sigma$ encompasses the entire Unicode range, which is common in modern programming languages, it consists of $\texttt{0x110000}$ distinct characters. Defining a transition from state $q$ to $q'$ that accepts any character in $\Sigma$ would require splitting into $\texttt{0x110000}$ individual transitions, rendering the product of two FAs highly inefficient. Furthermore, in practical applications, it is often necessary to consider infinite alphabets, such as the set of all integers.

Symbolic Finite Automata (SFA) and Symbolic Finite-state Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic models utilize transition labels defined by boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \% 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

Prior work has successfully formalized symbolic FAs within Isabelle/HOL through CertiStr \cite{cpp/KanLRS22}, demonstrating both the efficiency and effectiveness of symbolic FAs in practice. However, the formalization of FTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. FTs constitute a significantly more expressive and powerful theoretical framework compared to symbolic FAs, as evidenced by their capability to model complex string transformations such as replacement operations.
%
Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.

@@

In the refinement level, we implemented transition labels using an interval-based representation. The formalization of interval algebra provides efficient set-theoretic operations—including membership checking, intersection, and difference computations—facilitating the refinement of transition labels from abstract sets to concrete intervals. Furthermore, leveraging the data refinement framework \cite{DBLP:conf/itp/Lammich13} in Isabelle/HOL, we store states and transitions using sophisticated data structures such as hashmaps and red-black trees, ensuring efficient automata manipulation.

To evaluate the effectiveness of our formalization, we developed a string solver compliant with SMT-LIB language \cite{smtlib}, focusing particularly on replacement operations. We evaluated our implementation using a comprehensive set of benchmarks from the SMT-LIB repository \cite{smtlib_benchmarks}. The experimental results demonstrate that our formalization achieves computational efficiency in practical constraint-solving scenarios.

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite-state transducers in the Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop a comprehensive interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

The remainder of this paper is organized as follows:
Section \ref{sec:formalization} introduces the formal foundations of symbolic finite-state transducers.
Section \ref{sec:product-operation} presents the product operation at the abstract level.
Section \ref{sec_alg_refinement} details the algorithmic refinement of the product operation.
Section \ref{sec-app-str-solver} demonstrates the application to string constraint solving.
Section \ref{sec:related-work} discusses related work.
Section \ref{sec:conclusion} concludes with future directions.

@@

\emph{Symbolic Automata and Transducers.} Symbolic Automata and Transducers \cite{cav/DAntoniV17,VeanesHLMB12Transducer,popl/VeanesHLMB12, popl/DAntoniV14, entcs/DAntoniKW18, sofsem/TammV18} represent a significant advancement in automata theory, offering improved efficiency in operations and enhanced expressiveness through algebraic theories that support infinite alphabets. This symbolic framework has been progressively extended to accommodate more complex structures: symbolic tree transducers \cite{ershov/VeanesB11} handle hierarchical data structures, while symbolic pushdown automata \cite{cav/DAntoniA14} manage nested word structures. Recent developments have further expanded the scope of symbolic techniques to include B\"{u}chi automata and omega-regular languages \cite{pacmpl/VeanesBEZ25}, enabling verification of infinite-state systems and analysis of non-terminating computations.

\emph{Applications of Symbolic Automata and Transducers.} The efficiency, scalability, and expressive power of symbolic automata and transducers have led to their widespread adoption in numerous real-world applications. These applications span diverse domains, including: constraint solving for program analysis \cite{lpar/VeanesBM10}, security-critical sanitizer analysis for web applications \cite{uss/HooimeijerLMSV11}, runtime verification of system behaviors \cite{osdi/YaseenABCL20}, and automated program inversion for software transformation \cite{pldi/HuD17}. Each application leverages the symbolic approach's ability to handle complex patterns and infinite alphabets efficiently.

\emph{Formalization of Symbolic Automata and Transducers.} While classical automata theories have been extensively formalized in interactive theorem provers \cite{Tuerk-NFA, Lammich2014TheCA, Peter14, cpp/DoczkalKS13}, with some work on transducer formalization \cite{afp/LochmannFSTS21}, the symbolic variants of automata and transducers remain largely unexplored in formal verification. To our knowledge, CertiStr \cite{cpp/KanLRS22} represents the only existing work on symbolic automata formalization in a proof assistant. Our work advances this frontier by extending the formal treatment to both SFTs and $\varepsilon$SFAs.

\emph{String Solving.} A significant application of our certified transducer framework is in string constraint solving, a field that has seen intensive research development over the past decade. While our work provides formal verification guarantees, there exists a rich ecosystem of non-certified string solvers, each with distinct capabilities: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
As more and more bugs have beein uncovered in existing string solvers \cite{DBLP:conf/cav/BlotskyMBZKG18} and SMT solvers \cite{Mansur20} 
We believe that our work will benefit the community by providing a formal foundation for string solvers development.

\emph{Certified SMT Solvers.} Beyond string theories, certification efforts in SMT solving have extended to other domains. For example, the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of interactive theorem proving in certified SMT solver development.

@@

Future work includes extending our symbolic formalization in two key directions:
%
(1) Enriching the boolean algebra framework to support more complex theories, such as arithmetic constraints.
(2) Incorporating prioritized transitions into SFTs, which would enable precise modeling of SMT-LIB's standard semantics for replacement operations, particularly the first-match behavior.
These extensions will further enhance the expressiveness and practical applicability of our formalization while maintaining its formal verification guarantees.
